Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2015-03-11 papers/wdmerger_I/paper.tex
use IOP + set the paper size to letter so pdflatex makes things right
Insert description of the coupled rotation update. This used to be in the appendix and the content is now in the main rotation section. The method is implemented for rot_source_type == 4 in CASTRO.

In a rotating reference frame with angular frequency vector $\bm{\omega}$, the momentum equation becomes:
\begin{equation}
  \left.\frac{\partial(\rho \mathbf{u})}{\partial t}\right|_{\text{rot}} = -2\, {\bm\omega} \times (\rho\mathbf{u}) - \rho {\bm\omega} \times \left({\bm\omega} \times \mathbf{r}\right).
\end{equation}
Here $\mathbf{r}$ is the position vector with respect to the origin. Typically we choose $\bm{\omega} = (0, 0, 2\pi / T)^T$,
with the rotation axis coincident with the $z$ axis at $x = y = 0$.
$T$ is the rotation period, which is the most natural quantity to specify
for a rotating stellar system. As described in the appendix, we include this source term
in the edge state prediction in a way that is analogous to the gravity source.
We evaluate all quantities at cell centers and use the same predictor-corrector 
approach that we use for the gravity source terms to the momentum equations.

The update to the energy equation can be determined by taking the dot product of the velocity
with the momentum source terms. The Coriolis term vanishes identically, as it must
because the Coriolis term does no work on the fluid. The update from the centrifugal force becomes
\begin{equation}
  \left.\frac{\partial(\rho E)}{\partial t}\right|_{\text{rot}} = \frac{1}{\Delta V}\int \rho \mathbf{u} \cdot \mathbf{f}_R\, dV,
\end{equation}
with $\mathbf{f}_R \equiv  -{\bm\omega} \times \left({\bm\omega} \times \mathbf{r}\right)$. 
This expression is identical to the gravity source under the interchange of $\mathbf{g}$ with $\mathbf{f}_R$.
As observed by \cite{marcello:2012}, we can similarly write down a rotational potential,
\begin{equation}
  \Phi_R = -\frac{1}{2} \left| {\bm\omega} \times \mathbf{r} \right|^2.
\end{equation}
Therefore we can retrace the steps of Section \ref{sec:gravity_hydro_coupling} to obtain
an update to $(\rho E)$ that conserves the total energy in the presence of rotation,
\begin{equation}
  (\rho E_{\text{tot}}) = (\rho E) + \frac{1}{2} \rho \Phi + \rho \Phi_R.
\end{equation}
In one dimension, this would be
\begin{align}
  \left.\Delta(\rho E_i)\right|_{\text{rot}} &= \frac{1}{2} \Delta m_{i-1/2} (\Phi_{R_{i}} - \Phi_{R_{i-1}}) \notag \\
  &+ \frac{1}{2} \Delta m_{i+1/2} (\Phi_{R_{i+1}} - \Phi_{R_{i}}).
\end{align}
As with gravity, in three dimensions we include the mass fluxes through all six cell faces.

>>>>

In a rotating reference frame with angular frequency vector $\bm{\omega}$, the momentum equation becomes:
\begin{equation}
  \left.\frac{\partial(\rho \mathbf{u})}{\partial t}\right|_{\text{rot}} = -2\, {\bm\omega} \times (\rho\mathbf{u}) - \rho {\bm\omega} \times \left({\bm\omega} \times \mathbf{r}\right).
\end{equation}
Here $\mathbf{r}$ is the position vector with respect to the origin. Typically we choose $\bm{\omega} = (0, 0, 2\pi / T)^T$,
with the rotation axis coincident with the $z$ axis at $x = y = 0$.
$T$ is the rotation period, which is the most natural quantity to specify
for a rotating stellar system. As described in the appendix, we include this source term
in the edge state prediction in a way that is analogous to the gravity source.
We evaluate all quantities at cell centers. We use the same predictor-corrector 
approach that we use for the gravity source terms to the momentum equations. However, a slight 
difference is that the Coriolis force for each velocity component is coupled to other velocity 
components. If the rotation is about the $z$-axis, then the discrete update to 
$v_x^{n+1}$ depends on the value of $v_y^{n+1}$, and vice versa. If we fix the value of 
the time-level $n+1$ quantities after coming out of the hydrodynamics update, there 
would be a slight inconsistency between the $x$ and $y$ components of the velocity.

We propose a more accurate coupling that directly solves this implicit system of coupled 
equations. We denote by $(\widetilde{\rho \mathbf{u}})$ the value of the momentum after 
updating it with the centrifugal force, and the time-level $n$ Coriolis force. The remaining 
update for the time-level $n+1$ Coriolis force then appears as:
\begin{equation}
  (\rho \mathbf{u})^{n+1} = (\widetilde{\rho\mathbf{u}}) + \frac{\Delta t}{2} \left(-2\, {\bm\omega} \times (\rho\mathbf{u})^{n+1}\right)
\end{equation}
To proceed further, we assume that the rotation is about the $z$ axis with frequency $\omega$. 
Then there is no update to the $z$-momentum, and the other equations are:
\begin{align}
  (\rho u)^{n+1} &= (\widetilde{\rho u}) + \omega \Delta t (\rho v)^{n+1}) \\
  (\rho v)^{n+1} &= (\widetilde{\rho v}) - \omega \Delta t (\rho u)^{n+1})
\end{align}
We can directly solve this coupled system:
\begin{align}
  (\rho u)^{n+1} &= \frac{ (\widetilde{\rho u}) + \omega \Delta t (\widetilde{\rho v})}{1 + \omega^2 \Delta t^2} \\
  (\rho v)^{n+1} &= \frac{ (\widetilde{\rho v}) - \omega \Delta t (\widetilde{\rho u})}{1 + \omega^2 \Delta t^2}
\end{align}
We use this form of the momentum update in CASTRO. This improvement is small
but increases the accuracy of our rotating white dwarf systems over long time-scales.

The update to the energy equation can be determined by taking the dot product of the velocity
with the momentum source terms. The Coriolis term vanishes identically, and so
 the Coriolis term does no work on the fluid. The update from the centrifugal force becomes
\begin{equation}
  \left.\frac{\partial(\rho E)}{\partial t}\right|_{\text{rot}} = \frac{1}{\Delta V}\int \rho \mathbf{u} \cdot \mathbf{f}_R\, dV,
\end{equation}
with $\mathbf{f}_R \equiv  -{\bm\omega} \times \left({\bm\omega} \times \mathbf{r}\right)$. 
This expression is identical to the gravity source under the interchange of $\mathbf{g}$ with $\mathbf{f}_R$.
As observed by \cite{marcello:2012}, we can similarly write down a rotational potential,
\begin{equation}
  \Phi_R = -\frac{1}{2} \left| {\bm\omega} \times \mathbf{r} \right|^2.
\end{equation}
Therefore we can retrace the steps of Section \ref{sec:gravity_hydro_coupling} to obtain
an update to $(\rho E)$ that conserves the total energy in the presence of rotation,
\begin{equation}
  (\rho E_{\text{tot}}) = (\rho E) + \frac{1}{2} \rho \Phi + \rho \Phi_R.
\end{equation}
In one dimension, this would be
\begin{align}
  \left.\Delta(\rho E_i)\right|_{\text{rot}} &= \frac{1}{2} \Delta m_{i-1/2} (\Phi_{R_{i}} - \Phi_{R_{i-1}}) \notag \\
  &+ \frac{1}{2} \Delta m_{i+1/2} (\Phi_{R_{i+1}} - \Phi_{R_{i}}).
\end{align}
As with gravity, in three dimensions we include the mass fluxes through all six cell faces.
